---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# FIMSarch

<!-- badges: start -->
<!-- badges: end -->

The goal of FIMS architecture project is to compare statistical computing environments for use in the [FIMS](https://www.fisheries.noaa.gov/national/population-assessments/fisheries-integrated-modeling-system) intiative. This is ongoing research. For more details, see the [Working Document](https://andrea-havron-noaa.github.io/FIMSarch/docs/FIMS-architecture-exploratory-analysis.html)

## Resources
* [Julia GitHub](https://github.com/JuliaLang/julia)
* [Stan User's Guide](https://mc-stan.org/docs/2_19/stan-users-guide/)
* [Stan GitHub](https://github.com/stan-dev/stan)
* [cmdstan GitHub](https://github.com/stan-dev/stan)
* [TMB Documentation](https://kaskr.github.io/adcomp/_book/Introduction.html)
* [TMB GitHub](https://github.com/kaskr/adcomp)


## FIMS Architecture Organization
The FIMS architecture project is organized as follows:

```{r, toc, echo = FALSE, warning = FALSE}
library(pander)
library(kableExtra)
df <- data.frame(Directory = c("data",'docs','results','src','\\\ \\\ src/julia','\\\ \\\ src/R', '\\\ \\\ src/Rcpp', '\\\ \\\ src/stan', '\\\ \\\ src/tmb'),
                 Description = c("data simulation script, simulated data files, and initial values files",
                                 "working documents detailing project and results",
                                 "results files from benchmark study",
                                 "source files for project",
                                 "* stateSpace.jl: julia file that specifies and run gompertz and logistic models",
                                 "* model_setup.R: code used to setup TMB and Stan models
                                 \\\n * modular_demo.R: code runs modular example of passing C++ code to different platforms
                                 \\\n * run_models.R: runs different platforms for benchmark study excluding julia
                                 \\\n * utils.R: base functions for R code",
                                 "* Common.hpp:
                                 \\\n * likelihoods.hpp: likelihood definitions
                                 \\\n * logisticGrowth.cpp: TMB model specification
                                 \\\n * model.hpp: C++ model definitions",
                                 "* gompertz.stan: gompertz model
                                 \\\n * logistic.stan: logistic growth model",
                                 "* sptail_poisson.cpp: spatially explicit Poisson model
                                 \\\n * stateSpace.cpp: gompertz and logistic growth model"))
pander(df, keep.line.breaks = TRUE, style = 'grid', justify = 'left')
#kbl(df)
```
